---
title: "Formative Assessment 7: Part 2"
subtitle: "Applying Normal Distribution on Campus"
author: 
  - "Valdez, Chance Jayden"
  - "Tan, Heinz Nicole"
  - "Pinili, Nicklaus Vincent"
date: "`r Sys.Date()`"
output: 
  beamer_presentation:
    theme: "Madrid"
    colortheme: "default"
    fonttheme: "structurebold"
    slide_level: 2
---

## Background 

- The **objective** of this study is to measure and analyze Far Eastern University's facility utilization data using the occupancy proportion $\left(\frac{Occupied}{Capacity}\right)$.
- **Research Question:** Does campus facility occupancy rate $\left(\frac{Occupied}{Capacity}\right)$ follow a normal distribution, and what do its distribution characteristics reveal about campus space utilization?

## Methodology

```{r, echo=FALSE}
data <- read.csv("data.csv")

data$Proportion <- as.numeric(data$Proportion)
```

- The variable being studied is that of the occupancy proportion $\left(\frac{Occupied}{Capacity}\right)$.
- The gathered sample includes observation data from the FEU Pavilion from 10:00 to 10:30 and the second and third floors of the FEU Library from 12:00 to 1:00. The sample size consists of `r nrow(data)` instances of varying capacities ranging from a maximum of 12 tables to a minimum of 2 or 3.

```{r, echo=FALSE}
intervals <- cut(data$Proportion, 
                 breaks = seq(0, 1, by = 0.2), 
                 labels = c("0.01 - 0.20", 
                            "0.21 - 0.40", 
                            "0.41 - 0.60", 
                            "0.61 - 0.80", 
                            "0.81 - 1.00"),
                 include.lowest = TRUE)

table_clean <- as.data.frame(table(intervals))
colnames(table_clean) <- c("Interval", "Count")
```

## Frequency Distribution Table

The table below displays the frequency distribution for the calculated proportions across 143 observations. 

\begin{center}
```{r, echo=FALSE}
knitr::kable(table_clean, format = "latex", align = "cc")
```
\end{center}

## Mean and Standard Deviation

The mean and standard deviation are shown below:

- **Mean:** `r round(mean(data$Proportion, na.rm = TRUE), 4)`
- **Standard Deviation:** `r round(sd(data$Proportion, na.rm = TRUE), 4)`

## Normal Distribution Construction

The plot below displays the empirical distribution curve overlaid with the normal distribution curve.

```{r, echo=FALSE, fig.align='center', out.width='80%'}
proportions <- data$Proportion
mu <- mean(proportions, na.rm = TRUE)
sigma <- sd(proportions, na.rm = TRUE)

dens <- density(proportions, from = 0, to = 1)
plot(dens, 
     main = "Table Occupancy Data vs. dnorm() Curve", 
     xlab = "Proportion Value", 
     ylab = "Density", 
     xlim = c(-0.1, 1.1),
     ylim = c(0, max(dens$y) * 1.15))

curve(dnorm(x, mean = mu, sd = sigma), add = TRUE, col = "red")
```

## Percentage of Data Within 1, 2, and 3 Standard Deviations from the Mean

```{r, echo=FALSE}
pct_1s <- round(mean(abs(proportions - mu) <= 1 * sigma, na.rm = TRUE) * 100, 1)
pct_2s <- round(mean(abs(proportions - mu) <= 2 * sigma, na.rm = TRUE) * 100, 1)
pct_3s <- round(mean(abs(proportions - mu) <= 3 * sigma, na.rm = TRUE) * 100, 1)
```

Based on the occupancy proportion data gathered from the campus facilities, the actual distribution spreads across standard deviation thresholds as follows:

- **Within $\pm 1\sigma$ ($\mu \pm \sigma$):** `r pct_1s`% *(Theoretical: 68.3%)*
- **Within $\pm 2\sigma$ ($\mu \pm 2\sigma$):** `r pct_2s`% *(Theoretical: 95.4%)*
- **Within $\pm 3\sigma$ ($\mu \pm 3\sigma$):** `r pct_3s`% *(Theoretical: 99.7%)*

These empirical values can show how close the campus occupancy rate follows a theoretical normal distribution.

## Skewness, Outliers, and Shape
```{r, echo=FALSE}
n <- length(proportions)

skewness <- (n / ((n - 1) * (n - 2))) *
            sum(((proportions - mu) / sigma)^3)

Q1 <- quantile(proportions, 0.25, na.rm = TRUE)
Q3 <- quantile(proportions, 0.75, na.rm = TRUE)
IQR_value <- IQR(proportions, na.rm = TRUE)

lower_bound <- Q1 - 1.5 * IQR_value
upper_bound <- Q3 + 1.5 * IQR_value

outliers <- proportions[
  proportions < lower_bound |
  proportions > upper_bound
]

num_outliers <- length(outliers)
```

The distribution is approximately symmetric (skewness = `r round(skewness, 4)`), has `r num_outliers` outliers (bounds = `r round(lower_bound, 4)` to `r round(upper_bound, 4)`), and its shape suggests that occupancy proportions are relatively balanced around the mean but are not perfectly normally distributed.

## Findings
- The data shows that campus occupancy is generally moderate to high, with an approximate symmetric distribution but its not close to a normal distribution.
- These findings demonstrate that Pavilion and Library utilization varies across observations; thus, scheduling and space allocation can be adjusted for periods of higher occupancy (such as exam seasons) or lower occupancy (such as synchronous learning days).
- The data can help the university monitor facility utilization and make informed decisions about campus space management.
- To optimize campus space, the university should implement real-time occupancy tracking and dynamic scheduling that shifts resources during peak exam seasons and utilizes low-occupancy synchronous learning days for maintenance.

## Summary 

- **Sample size:** `r n`
- **Mean (\(\mu\)):** `r round(mu, 4)`
- **Standard deviation (\(\sigma\)):** `r round(sigma, 4)`
- **Skewness:** `r round(skewness, 4)`
- **\(Q_1\):** `r round(Q1, 4)`
- **\(Q_3\):** `r round(Q3, 4)`
- **IQR:** `r round(IQR_value, 4)`
- **Outlier Bounds:** `r round(lower_bound, 4)` to `r round(upper_bound, 4)`
- **Number of Outliers:** `r num_outliers`
- **Within \(\pm1\sigma\):** `r round(pct_1s, 1)`%
- **Within \(\pm2\sigma\):** `r round(pct_2s, 1)`%
- **Within \(\pm3\sigma\):** `r round(pct_3s, 1)`%
- Analysis codes can be found in the .Rmd file!